---
output:
  pdf_document: default
  html_document: default
---
# Replication

In this notebook, we replicate the **Impacts on Target Outcomes**
section of the chosen paper. We restrict our view to the key outcomes of
rates of attendance, truancy, and suspension, along with academic achievement
as measured by reading and math scores on standardized examinations.

```{r, message=F}
# data loading utilities
source("load.R")
```

## Randomization Check

Load in baseline covariates about the schools and random assignments.
```{r}
school_df <- load_school_info()

baseline_covariates <- c(
  "FRL", "PUPTCH", "PCT_BL", "PCT_HS", "PCT_WH", 
  "SMALLSCHOOL", "LARGESCHOOL", "GRDRNG_ELEM", "GRDRNG_HS"
  )

```

```{r}
library(modelsummary)

to_display <- school_df %>% 
  select(all_of(baseline_covariates)) %>%
  rename(
      "Proportion Eligible for Free/Reduced Lunch" = FRL,
      "Student-Teacher Ratio" = PUPTCH,
      "% Black" = PCT_BL,
      "% Hispanic" = PCT_HS,
      "% White" = PCT_WH,
      "Small 6th and 7th Grades (less than 200 students)" = SMALLSCHOOL,
      "Large 6th and 7th Grades (more than 700 students)" = LARGESCHOOL,
      "Also an Elementary School" = GRDRNG_ELEM,
      "Also a High School" = GRDRNG_HS,
  )

# Create the table object
make_tbl <- function(output) {
  datasummary_balance(
    ~ 1,
    data = to_display,
    title = "Summary Statistics",
    fmt = 3,   # number of decimal places
    # notes = "Note: Standard deviations in parentheses.",
    output = output
  )
}

make_tbl("kableExtra")
# make_tbl("doc/baseline_covariates.tex")
```

We'll perform a block balanced randomization check.

```{r}
baseline_covariates <- c(
  "FRL", "PUPTCH", "PCT_BL", "PCT_HS", "PCT_WH", 
  "SMALLSCHOOL", "LARGESCHOOL", "GRDRNG_ELEM", "GRDRNG_HS"
  )
formula <- as.formula(paste(
  "TREATMENT ~", paste(baseline_covariates, collapse=" + "), " + factor(COHORT)"
  ))

g <- lm(formula, data=school_df)

f_stat <- summary(g)$fstatistic
p_val <- pf(f_stat["value"], f_stat["numdf"], f_stat["dendf"], lower.tail=FALSE)

print(sprintf(
    "Joint F-test of covariate balance: F(%d, %d) = %.3f (p = %.3f)",
    f_stat["numdf"], 
    f_stat["dendf"], 
    f_stat["value"],
    p_val
))
```

Great! We can be reasonably confident that the randomization procedure was
performed correctly then, and that there are no strange inaccuracies from the
matching process.

## School-Level: Attendance, Truancy, and Suspension

Load in the data.
```{r}
# attendance, truancy, suspensions
school_level_outcomes <- load_school_outcomes()
```

View a few random rows for clarity.

\emph{Attendance} is measured as the total number of student days attended, divided by
the total number of days school was in session. It's unclear exactly what this
means, but we'll assume that this is the average attendance rate across all
students in the school since all values are $< 1$.

\emph{Truancy} is defined as the total number of student unexcused absences, divided by
the total number of days school was in session. Again unclear, but we'll assume
that this is the average truancy rate across all students in the school.

\emph{Suspension rates} are defined as the number of suspensions divided by the
total number of students in the school.

```{r}
set.seed(0527)
# Rename columns for compact display:
col_mapping <- c(
  COHORT = "COHORT",
  Z = "TREATMENT",
  PID = "PID",
  # Attendance years 0-2
  A0 = "ATTEND_Y0",
  A1 = "ATTEND_Y1", 
  A2 = "ATTEND_Y2",
  # Truancy years 0-2
  T0 = "TRUANCY_Y0",
  T1 = "TRUANCY_Y1",
  T2 = "TRUANCY_Y2",
  # Suspensions years 1-2
  S_IN1 = "SUS_IN_SCH_YR1",
  S_IN2 = "SUS_IN_SCH_YR2",
  S_OUT1 = "SUS_OUT_SCH_YR1",
  S_OUT2 = "SUS_OUT_SCH_YR2"
)
df_to_dply <- school_level_outcomes %>%
  select(all_of(unname(col_mapping))) %>%
  rename(!!!col_mapping)

knitr::kable(
  df_to_dply %>% sample_n(5),
  digits=3,
  caption="School-Level Outcome Information"
  )
```

### Attendance
First, we replicate their analysis on attendance. The authors run two separate
models for each outcome variable, one for 1-year effects and one for 2-year
effects.
```{r}
attendance <- load_school_level_outcome("attendance")

formula <- as.formula(paste(
  "ATTEND_Y1 ~",
  "TREATMENT + ATTEND_Y0 + ",
  # postpend _Y1 since each covariate is measured in both years
  paste(paste0(baseline_covariates, "_Y1"), collapse=' + '),
  "+",
  paste(paste0('PAIR', 2:24), collapse=" + ")
))

g1 <- lm(formula, attendance)

summary(g1)
```
Nothing significant achieved, matching the paper. We do obtain a different
coefficient value, however, interestingly. Our $\beta_1 = 0.006$ and in the
paper they obtain $0.01$, though this could be due to rounding to two
significant figures in the display.

```{r}
formula <- as.formula(paste(
  "ATTEND_Y2 ~",
  "TREATMENT + ATTEND_Y0 + ",
  # postpend _Y1 since each covariate is measured in both years
  paste(paste0(baseline_covariates, "_Y1"), collapse=' + '),
  "+",
  paste(paste0('PAIR', 2:24), collapse=" + ")
))

g2 <- lm(formula, attendance)

summary(g2)
```
Again, we match the paper and find no significant effects.

### Truancy

```{r}
truancy <- load_school_level_outcome("truancy")

formula <- as.formula(paste(
  "TRUANCY_Y1 ~",
  "TREATMENT + TRUANCY_Y0 + ",
  # postpend _Y1 since each covariate is measured in both years
  paste(paste0(baseline_covariates, "_Y1"), collapse=' + '),
  "+",
  paste(paste0('PAIR', 2:24), collapse=" + ")
))

g1 <- lm(formula, truancy)

summary(g1)
```

```{r}
truancy <- load_school_level_outcome("truancy")

formula <- as.formula(paste(
  "TRUANCY_Y2 ~",
  "TREATMENT + TRUANCY_Y0 + ",
  # postpend _Y1 since each covariate is measured in both years
  paste(paste0(baseline_covariates, "_Y1"), collapse=' + '),
  "+",
  paste(paste0('PAIR', 2:24), collapse=" + ")
))

g2 <- lm(formula, truancy)

summary(g2)
```
### Curiosity: Correlation between Truancy and Attendance

```{r}
par(mfrow=c(1, 2))
xlim <- c(0.8, 1.0)
ylim <- c(0.0, 0.1)

plot(
  school_level_outcomes$ATTEND_Y1,
  school_level_outcomes$TRUANCY_Y1,
  xlab='Attendance, Year 1',
  ylab='Truancy, Year 1',
  xlim=xlim, ylim=ylim
  )
plot(
  school_level_outcomes$ATTEND_Y2,
  school_level_outcomes$TRUANCY_Y2,
  xlab='Attendance, Year 2',
  ylab='Truancy, Year 2',
  xlim=xlim, ylim=ylim
  )

mtext(
  "Correlation between Attendance and Truancy Outcomes",
      side = 3,
      line = - 2,
      outer = TRUE
      )
```


## Individual-Level: Achievement

```{r}
achievement <- load_achievement()
```

The researchers employed a matched-pairs experimental design with a waitlist
control structure, divided into two cohorts. Cohort 1 data was measured during the 2016-17
and 2017-18 school years; Cohort 2 started one year later, with outcome data measured
therefore during the 2017-18 and 2018-19 school years.

Observations were deleted "listwise" when outcome data were missing. No analysis of
this attrition was performed in the paper.

```{r}
# col_mapping <- c(
#   Z="TREATMENT",
#   GRADE="GRADE",
#   SCHOOL_YEAR="SCHOOL_YEAR",
#   paste0(
#     "Y",
#     rep(1:2, each=2),
#     c("BASELINE_R_Z", "BASELINE_M_Z")
#   )
# )
knitr::kable(
  achievement %>% sample_n(5),
  digits=3,
  caption="School-Level Outcome Information"
  )
```
```{r}
achievement %>% group_by(PID)
```
The researchers estimated two separate sets of models, one for reading scores and
one for math scores.

### Year 1 Models

```{r}
library(lme4)
library(lmerTest)

# In this regression, the authors do NOT consider grade range of the school.
# It is unclear why not.
indreg_school_covariates <- c(
  "FRL", "PUPTCH", "PCT_BL", "PCT_HS", "PCT_WH", 
  "SMALLSCHOOL", "LARGESCHOOL"
)

ind_covariates <- c("FEMALE", "WHITE")

year_1_grade_7_df <- achievement %>%
  filter(!is.na(Y1BASELINE_R_Z), !is.na(Y1BASELINE_M_Z), GRADE=="07") %>%
  mutate(
    Y_R=ZSCORE_R07,
    Y_M=ZSCORE_M07
    )

formula <- as.formula(paste(
  "Y_R ~",
  "TREATMENT + Y1BASELINE_R_Z + ",
  paste(ind_covariates, collapse=" + "),
  " + ",
  paste(paste0('PAIR', 2:24), collapse=" + "),
  "+",
  paste(paste0("(1|", indreg_school_covariates, ")"), collapse=' + ')
))

g1 <- lmer(formula, year_1_grade_7_df)
summary(g1)
```
```{r}
formula <- as.formula(paste(
  "Y_M ~",
  "TREATMENT + Y1BASELINE_M_Z + ",
  paste(ind_covariates, collapse=" + "),
  " + ",
  paste(paste0('PAIR', 2:24), collapse=" + "),
  "+",
  paste(paste0("(1|", indreg_school_covariates, ")"), collapse=' + ')
))

g1_m_7 <- lmer(formula, year_1_grade_7_df)
summary(g1_m_7)
```

```{r}
year_1_grade_8_df <- achievement %>%
  filter(!is.na(Y1BASELINE_R_Z), !is.na(Y1BASELINE_M_Z), GRADE=="08") %>%
  mutate(
    Y_R=ZSCORE_R08,
    Y_M=ZSCORE_M08
    )

formula <- as.formula(paste(
  "Y_R ~",
  "TREATMENT + Y1BASELINE_R_Z + ",
  paste(ind_covariates, collapse=" + "),
  " + ",
  # postpend _Y1 since each covariate is measured in both years
  paste(paste0('PAIR', 2:24), collapse=" + "),
  "+",
  paste(paste0("(1|", indreg_school_covariates, ")"), collapse=' + ')
))

g1_r_8 <- lmer(formula, year_1_grade_8_df)
summary(g1_r_8)
```
```{r}
formula <- as.formula(paste(
  "Y_M ~",
  "TREATMENT + Y1BASELINE_M_Z + ",
  # postpend _Y1 since each covariate is measured in both years
  paste(ind_covariates, collapse=" + "),
  " + ",
  paste(paste0('PAIR', 2:24), collapse=" + "),
  "+",
  paste(paste0("(1|", indreg_school_covariates, ")"), collapse=' + ')
))

g1_m_8 <- lmer(formula, year_1_grade_8_df)
summary(g1_m_8)
```
### Year 2 Models

```{r}
schoolsize <- achievement %>%
  filter(!is.na(LARGESCHOOL)) %>%
  group_by(PID, TREATMENT) %>%
  select(PID, TREATMENT, LARGESCHOOL) %>%
  distinct()

year_2_df <- achievement %>%
  filter(!is.na(Y2BASELINE_R_Z), !is.na(Y2BASELINE_M_Z)) %>%
  # only 8th graders in the sample
  mutate(
    Y_R=ZSCORE_R08,
    Y_M=ZSCORE_M08
    ) %>%
  left_join(
    schoolsize, by=c("PID", "TREATMENT")
  ) %>%
  select(-LARGESCHOOL.x) %>%  # remove the old NA column 
  rename(LARGESCHOOL = LARGESCHOOL.y)  # rename the new column

formula <- as.formula(paste(
  "Y_R ~",
  "TREATMENT + Y2BASELINE_R_Z + ",
  paste(ind_covariates, collapse=" + "),
  " + ",
  paste(paste0('PAIR', 2:24), collapse=" + "),
  "+",
  paste(paste0("(1|", indreg_school_covariates, ")"), collapse=' + ')
))

g2_r_8 <- lmer(formula, year_2_df)
summary(g2_r_8)
```
```{r}
formula <- as.formula(paste(
  "Y_M ~",
  "TREATMENT + Y2BASELINE_M_Z + ",
  paste(ind_covariates, collapse=" + "),
  " + ",
  paste(paste0('PAIR', 2:24), collapse=" + "),
  "+",
  paste(paste0("(1|", indreg_school_covariates, ")"), collapse=' + ')
))

g2_m_8 <- lmer(formula, year_2_df)
summary(g2_m_8)
```

## Robustness to Model Specification

In this section, we probe the analysis's robustness to model specification
by grouping the grade analyses. We also hope to find alternative ways of analyzing
the data to improve parameter standard errors.

We'll try to estimate 1-year grade-specific effects together in one model,
rather than two separate models.
```{r}
year_1_df <- achievement %>%
  filter(!is.na(Y1BASELINE_R_Z), !is.na(Y1BASELINE_M_Z)) %>%
  mutate(
    Y_R=if_else(
      GRADE=="07",
      ZSCORE_R07,
      ZSCORE_R08
    ),
    Y_M=if_else(
      GRADE=="07",
      ZSCORE_M07,
      ZSCORE_M08
    ),
    GRADE_8 = (GRADE == "08")
  )

formula <- as.formula(paste(
  "Y_R ~",
  "TREATMENT:GRADE_8 + Y1BASELINE_R_Z + ",
  paste(ind_covariates, collapse=" + "),
  " + ",
  paste(paste0('PAIR', 2:24), collapse=" + "),
  "+",
  paste(paste0("(1|", indreg_school_covariates, ")"), collapse=' + ')
))

g1 <- lmer(formula, year_1_df)
summary(g1)
```


```{r}
formula <- as.formula(paste(
  "Y_M ~",
  "TREATMENT:GRADE_8 + Y1BASELINE_M_Z + ",
  paste(ind_covariates, collapse=" + "),
  " + ",
  paste(paste0('PAIR', 2:24), collapse=" + "),
  "+",
  paste(paste0("(1|", indreg_school_covariates, ")"), collapse=' + ')
))

g1_m <- lmer(formula, year_1_df)
summary(g1_m)
```

